Variation in site entropy explains differences in structure-sequence relationships of proteins

### 2015-06-03 manuscript/manuscript.tex
Addressed Amir's comments

\begin{abstract}
Recent work has shown that structural properties are capable of predicting site-specific sequence variability for a given protein. However, the strength and significance of these structure-sequence relations appear to vary widely among different proteins, with absolute correlation strengths ranging from $0.1$ to $0.8$.  Additionally, there have been conflicting reports about about whether local packing density or solvent accessibility is the best predictor of site variability in proteins. According to \cite{Yehetal2014}, local packing density is the strongest predictor of site-wise variability. However, recently, \cite{Shahmoradietal2014} compared local packing density and solvent accessibility in viral proteins and found that relative solvent accessibility is a stronger predictor of site-wise variability. Here we present research that suggests that differences in the correlations between these two datasets are due to differences in the site variability among proteins within in each dataset. Specifically proteins with a larger variance in entropy among sites exhibit stronger structure-sequence correlations between both local packing density and solvent accessibility. 
\end{abstract}
\vfill
\vfill
\def\thefootnote{\fnsymbol{footnote}}
\setcounter{footnote}{0}

@@

Proteins are subject to a number of biophysical and functional constraints \citep{Scherreretal2012, Wilkeetal2010}. These constraints result in site-specific patterns of sequence variability within a protein. Recently several site-specific structural properties that can explain patterns of sequence variability in proteins have been identified. One of the earliest examples was Relative Site Accessibility (RSA). \cite{Fransozaetal2009} identified RSA as the strongest predictor of evolutionary rate and found that residues that are buried within the core of proteins tend to be more conserved than exposed residues close to the surface of the protein. In their analysis, they considered the ability of both RSA and various definitions of residue packing density to predict evolutionary rate. They found that RSA and evolutionary rate shared a significant linear relationship. Afterwards, several other works \citep{Ramseyetal2011, Scherreretal2012} also found that RSA as a significant predictor of evolutionary rate and supported this linear relationship. However, these papers all have the same flaw. During the course of their analysis they binned the protein sites and averaged over all sites within a bin when determining the trend of RSA. This process may have produced artifacts that account for this strong linear trend between RSA and evolutionary rate. \\
\indent Recently, \cite{Yehetal2014} performed a similar analysis on a series of enzyme monomer proteins and found that packing density, as defined by Contact Number and Weighted Contact Number \citep{Liaoetal2005, Yehetal2014, Huangetal2014}, was the strongest determinant of site variability.  Soon Afterwards, \cite{Shahmoradietal2014} also performed a site-wise analysis on a series of viral proteins. In this analysis they found that RSA had the strongest correlation with site variability as opposed to local packing density. Additionally the effect seen between CN and WCN was of a much smaller magnitude as compared to \cite{Yehetal2014}. Therefore the relationship between local pacing density, RSA and site-specific measures of sequence variability is not well understood. Here we attempt to reconcile the work done in this area. We find that site variability is the primary determinant of the strength of structure-sequence relationships and some differences in previous work can be explained in terms of differing levels of site variability. \\

@@

\indent  The average site entropy of a given protein does not seem to be a significant determinant of the strength of structural correlations (Figures \ref{fig:entropy_cn_cor}A, \ref{fig:entropy_wcn_cor}A, \ref{fig:entropy_rsa_cor}A). However, when examining the variance of entropy there is a clear trend within the enzyme proteins.  Proteins with a higher variance in site variability across the protein typically have lower correlations (Figures \ref{fig:entropy_cn_cor}B, \ref{fig:entropy_wcn_cor}B, \ref{fig:entropy_rsa_cor}B). If you extrapolate the trend from the enzyme proteins, the viral proteins follow a similar trend. The correlation between the $\rho$ between RSA and entropy and the variance of entropy is positive. Proteins with a larger variance in site entropy have the strongest correlations. The correlation between local packing density (both WCN and CN) and the variance of entropy is negative. Unlike entropy, there is no relationship between the variance of evolutionary rates at sites and any of the measured structural properties (Figures \ref{fig:var_rate_cor}, \ref{fig:var_rate_rsa_cor}). There also is a wide spread in the variance of the evolutionary rates across proteins for both the enzyme and viral proteins.

@@

\indent In order to further examine the relationship between entropy and the structure-sequence relationships, we used the mean entropy and variance of entropy at sites as predictors of structure-sequence relationships. Table \ref{table:model_stats} illustrates the coefficients of various linear models.  For WCN-Entropy correlations, mean entropy is not a significant predictor. Although for CN-Entropy correlations, mean entropy is significant the coefficient is 0.079 and therefore predictive power is low. Therefore the mean site variability of a protein as measured by entropy is not indicative of whether packing density can be using to predict site variation.  Mean Entropy is also not a significant predictor for the strength of the relationship between RSA and Entropy. For all linear models where mean entropy and dataset were used to predict structure-sequence correlations, dataset was a significant predictor (Table \ref{table:model_stats}). This means that enzyme and virus proteins have different correlations when using mean entropy as a predictor. This agrees with differences seen in the previous works by \cite{Yehetal2014} and \cite{Shahmoradietal2014}.

\indent However, when looking at the variance in entropy there are some stark differences. For all structural predictors (i.e., CN, WCN and RSA), the variance in entropy at sites within a protein is a significant predictor of the strength of structure-sequence relations. For packing density, proteins with a larger variance in site-wise entropy have lower correlations. Proteins with a higher variance in entropy tend to have stronger RSA-Entropy correlations. When using variance and dataset as predictors of structure-sequence correlations, dataset was not a significant predictor in any model. When looking at the variance of entropy, proteins within both datasets at in a similar fashion and overall trends seen between the variance of site variation and the strength of structure-sequence relationships is preserved across both datasets.

\section{Discussion}
\label{sec:dcr}
Structure imposes constraints on the evolution of proteins. Therefore structural predictors can be used to predict the variability at sites. Two main determinants of variability at protein sites are solvent accessibility and local packing density. Sites that are on surface of the protein tend to have higher solvent accessibility and exhibit more site variability. Sites that are densely packed and have more contacts tend to evolve slower and exhibit less sequence variability. However, it has not been well understood which of these two predictors, local packing density and solvent accessibility, is the main determinant of site variability in proteins. Two recent papers \citep{Shahmoradietal2014, Yehetal2014} give conflicting reports on which is the best predictor. In a study done on monomeric enzyme proteins, \cite{Yehetal2014} reported that local packing density, as measured by Contact Number and Weighted Contact Number, is the strongest determinant of site variability. \cite{Shahmoradietal2014} performed a similar analysis on a series of viral proteins and found that solvent accessibility is more predictive of site variability.

\indent Here we examined the relationship between site variability and the strength of structure-sequence relationships. We found that proteins with a larger variance in site variability as measured by sequence entropy, have stronger structure-sequence relationships on average. However, when looking at the mean entropy values of proteins there is a distinct difference between the virus and enzyme datasets. Overall the enzyme proteins have more site variability than the viral proteins.When predicting structure-sequence relationships the variance in site variability among proteins is a better predictor than mean site variability.  Additionally the reason that the enzyme proteins in \cite{Yehetal2014} had higher correlation coefficients is that on average, the enzyme proteins have larger variance in entropy as compared to the virus proteins of \cite{Shahmoradietal2014}. Moreover, since the variance in entropy is a more significant predictor of correlations in local packing density, an increase level of variance in site variability will have a larger effect on the local packing density correlations. This would explain why the local packing density correlations were of a higher magnitude compared to the RSA correlations within the enzyme proteins of \cite{Yehetal2014}. It appears that site entropy and evolutionary rate are not equal measures of site variability in terms of predicting the strength of structure-sequence relationships. Finally, the variance in evolutionary rate among sites of a protein does not have an effect on the correlation magnitude. This suggests that the absolute variability at sites is a better predictor of whether structural determinants can be used to predict variability at sites  as opposed to evolutionary rate. Therefore when examining structural determinants of site variability, entropy is a may be a more meaningful quantity.

>>>>

\begin{abstract}
Recent work has shown that structural properties are capable of predicting site-specific sequence variability for a given protein. However, the strength and significance of these structure-sequence relations appear to vary widely among different proteins, with absolute correlation strengths ranging from $0.1$ to $0.8$. There have two works that have investigated structural predictors of site variability which both present different results based on the strength of correlations of structural predictors. According to \cite{Yehetal2014}, local packing density is the strongest predictor of site-wise variability. However, recently, \cite{Shahmoradietal2014} compared local packing density and solvent accessibility in viral proteins and found that relative solvent accessibility is a stronger predictor of site-wise variability. Here we present research that suggests that differences in the correlations between these two datasets are due to differences in the site variability among proteins within in each dataset. Specifically proteins with a larger variance in entropy among sites exhibit stronger structure-sequence correlations between both local packing density and solvent accessibility. 
\end{abstract}
\vfill
\vfill
\def\thefootnote{\fnsymbol{footnote}}
\setcounter{footnote}{0}

@@

Proteins are subject to a number of biophysical and functional constraints \citep{Scherreretal2012, Wilkeetal2010}. These constraints result in site-specific patterns of sequence variability within a protein. Recently several site-specific structural properties that can explain patterns of sequence variability in proteins have been identified. One of the earliest examples was Relative Site Accessibility (RSA). \cite{Fransozaetal2009} identified RSA as the strongest predictor of evolutionary rate and found that residues that are buried within the core of proteins tend to be more conserved than exposed residues close to the surface of the protein. In their analysis, they considered the ability of both RSA and various definitions of residue packing density to predict evolutionary rate. They found that RSA and evolutionary rate shared a significant linear relationship. Afterwards, several other works  also found that RSA as a significant predictor of evolutionary rate and supported this linear relationship \citep{Ramseyetal2011, Scherreretal2012}. However, these papers all have the same flaw. During the course of their analysis they binned the protein sites and averaged over all sites within a bin when determining the trend of RSA. This process may have produced artifacts that account for this strong linear trend between RSA and evolutionary rate. \\
\indent Recently, \cite{Yehetal2014} performed a similar analysis on a series of enzyme monomer proteins and found that packing density, as defined by Contact Number and Weighted Contact Number \citep{Liaoetal2005, Yehetal2014, Huangetal2014}, was the strongest determinant of site variability.  Soon Afterwards, \cite{Shahmoradietal2014} also performed a site-wise analysis on a series of viral proteins.{\color{blue} In this analysis they found that RSA had the strongest correlation with site variability as opposed to local packing density.} Additionally the effect seen between CN and WCN was of a much smaller magnitude as compared to \cite{Yehetal2014}. Therefore the relationship between local pacing density, RSA and site-specific measures of sequence variability is not well understood. Here we attempt to reconcile the work done in this area. We find that site variability is the primary determinant of the strength of structure-sequence relationships and some differences in previous work can be explained in terms of differing levels of site variability. \\

@@

\indent  The average site entropy of a given protein does not seem to be a significant determinant of the strength of structural correlations (Figures \ref{fig:entropy_cn_cor}A, \ref{fig:entropy_wcn_cor}A, \ref{fig:entropy_rsa_cor}A). However, when examining the variance of entropy there is a clear trend within the enzyme proteins.  Proteins with a higher variance in site variability across the protein typically higher  correlations in magnitude (Figures \ref{fig:entropy_cn_cor}B, \ref{fig:entropy_wcn_cor}B, \ref{fig:entropy_rsa_cor}B). If you extrapolate the trend from the enzyme proteins, the viral proteins follow a similar trend. The correlation between the $\rho$ between RSA and entropy and the variance of entropy is positive. Proteins with a larger variance in site entropy have the strongest correlations. The correlation between local packing density (both WCN and CN) and the variance of entropy is negative. Unlike entropy, there is no relationship between the variance of evolutionary rates at sites and any of the measured structural properties (Figures \ref{fig:var_rate_cor}, \ref{fig:var_rate_rsa_cor}). There also is a wide spread in the variance of the evolutionary rates across proteins for both the enzyme and viral proteins.

@@

\indent In order to further examine the relationship between entropy and the structure-sequence relationships, we used the mean entropy and variance of entropy at sites as predictors of the strength of structure-sequence relationships. Table \ref{table:model_stats} illustrates the coefficients of various linear models.  For WCN-Entropy correlations, mean entropy is not a significant predictor. Although for CN-Entropy correlations, mean entropy is significant the coefficient is 0.079 and therefore predictive power is low. Mean Entropy is also not a significant predictor for the strength of the relationship between RSA and Entropy. For all linear models where mean entropy and dataset were used to predict structure-sequence correlations, dataset was a significant predictor (Table \ref{table:model_stats}). This means that enzyme and virus proteins have different correlations when using mean entropy as a predictor. This agrees with differences seen in the previous works by \cite{Yehetal2014} and \cite{Shahmoradietal2014}.

\indent However, when looking at the variance in entropy there are some stark differences. For all structural predictors (i.e., CN, WCN and RSA), the variance in entropy at sites within a protein is a significant predictor of the strength of structure-sequence relations. For packing density, proteins with a larger variance in site-wise entropy have more negative correlations that are higher in magnitude. Proteins with a higher variance in entropy tend to have stronger RSA-Entropy correlations. When using variance and dataset as predictors of structure-sequence correlations, dataset was not a significant predictor in any model. When looking at the variance of entropy, proteins within both datasets at in a similar fashion and overall trends seen between the variance of site variation and the strength of structure-sequence relationships is preserved across both datasets.

\section{Discussion}
\label{sec:dcr}
Structure imposes constraints on the evolution of proteins. Therefore structural predictors can be used to predict the variability at sites. Two main determinants of variability at protein sites are solvent accessibility and local packing density. Sites that are on surface of the protein tend to have higher solvent accessibility and exhibit more site variability. Sites that are densely packed and have more contacts tend to evolve slower and exhibit less sequence variability. {\color{blue} However, it has not been well understood which of these two predictors, local packing density and solvent accessibility, is the main determinant of site variability in proteins. Two recent papers \citep{Shahmoradietal2014, Yehetal2014} give conflicting reports on which is the best predictor. In a study done on monomeric enzyme proteins, \cite{Yehetal2014} reported that local packing density, as measured by Contact Number and Weighted Contact Number, is the strongest determinant of site variability. \cite{Shahmoradietal2014} performed a similar analysis on a series of viral proteins and found that solvent accessibility is more predictive of site variability. AMIR THINKS THIS IS NOT RELEVANT!}

\indent Here we examined the relationship between site variability and the strength of structure-sequence relationships. We found that proteins with a larger variance in site variability as measured by sequence entropy, have stronger structure-sequence relationships on average. However, when looking at the mean entropy values of proteins there is a distinct difference between the virus and enzyme datasets. Overall the enzyme proteins have more site variability than the viral proteins.When predicting structure-sequence relationships the variance in site variability among proteins is a better predictor than mean site variability.  Additionally the reason that the enzyme proteins in \cite{Yehetal2014} had higher correlation coefficients is that on average, the enzyme proteins have larger variance in entropy as compared to the virus proteins of \cite{Shahmoradietal2014}. Moreover, since the variance in entropy is a more significant predictor of correlations in local packing density, an increased level of variance in site variability will have a larger effect on the local packing density correlations. {\color{blue} This would explain why the local packing density correlations were of a higher magnitude compared to the RSA correlations within the enzyme proteins of \cite{Yehetal2014}. THIS NEEDS TO BE EXPLAINED}. It appears that site entropy and evolutionary rate are not equal measures of site variability in terms of predicting the strength of structure-sequence relationships. Finally, the variance in evolutionary rate among sites of a protein has a smaller an effect on the correlation magnitude as compared to entropy. This suggests that the absolute variability at sites is a better predictor of whether structural determinants can be used to predict variability at sites  as opposed to evolutionary rate. Therefore when examining structural determinants of site variability, entropy is a may be a more meaningful quantity.

{\color{red} Amir's comment: I think it maybe better to not put so much emphasis on variance of entropy, but its implications and meaning. For example entropy variance is an indicator of sequence divergence, to some extent.}
